ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-09-26 paper/aaai/author_response.qmd
tables updated

@@

One reviewer complained about the mathematical notation, a concern that was not shared by any of the other reviewers. Nonetheless, we have revisited the notation and hope that it is now more clear. That same reviewer also raised concern about our definitions of plausibility and faithfulness that rely on distributional properties. We have extensively argued our case during the rebuttal and pointed to a potential reviwer misunderstanding in this context. None of the other reviewers found any issue with our definitions and we have made no changes in this regard. We did, however, make a minor change with respect to the related evaluation metrics. We are now more careful about our choice of the distance function. In particular, we investigated various distance metrics for image data and decided to rely on structural dissimilarity. For all other data we use the L2 Norm, where we previously used the L1 Norm. This has no impact on the results, but there was no obvious reason to use the L1 Norm in the first place other than the fact that it is typically used to assess closeness.

>>>>

## Beyond JEMs

Another common concern was that *ECCCo* primiarly achieved good results for JEMs.

@@

One reviewer took issue with our mathematical notation, a concern that was not shared by any of the other reviewers. Nonetheless, we have revisited the notation and hope that it is now more clear. That same reviewer also raised concern about our definitions of plausibility and faithfulness that rely on distributional properties. We have extensively argued our case during the rebuttal and pointed to a potential reviwer misunderstanding in this context. None of the other reviewers found any issue with our definitions and we have made no changes in this regard. We did, however, make a minor change with respect to the related evaluation metrics. We are now more careful about our choice of the distance function. In particular, we investigated various distance metrics for image data and decided to rely on structural dissimilarity. For all other data we use the L2 Norm, where we previously used the L1 Norm. This has no impact on the results, but there was no obvious reason to use the L1 Norm in the first place other than the fact that it is typically used to assess closeness.

### 2023-09-26 paper/tables.Rmd
Cynthia's comments addressed
tables updated

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i[,variable:=factor(variable, levels=measures)]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,model:=factor(model,levels=chosen_model)]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep, table.env="table*"
) %>%
  kable_styling(
    latex_options = c("scale_down")
  ) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

@@

```{r}
for (name in unique(tab$dataname)) {
  data_indicator <- gsub(" ", "-", tolower(unique(name)))
  # Choices:
  tab_full <- dcast(tab[dataname==name], model + generator ~ variable)
  col_names <- c(
    "Model",
    "Generator",
    measure_names
  )
  algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
  file_name <- sprintf(
    "paper/contents/table-%s.tex", 
    data_indicator
  )
  cap <- sprintf(
    "All results for %s dataset: sample averages +/- one standard deviation over all counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
    name,
    data_indicator
  )
  kbl(
    tab_full, caption = cap, 
    align = "c", col.names=col_names, booktabs = T, escape=F, 
    format="latex"
  ) %>%
    kable_styling(latex_options = c("scale_down")) %>%
    kable_paper(full_width = F) %>%
    collapse_rows(columns = 1:2, latex_hline = "custom", valign = "top") %>%
    save_kable(file_name)
}

@@

```{r}
for (name in unique(tab_valid$dataname)) {
  data_indicator <- gsub(" ", "-", tolower(unique(name)))
  # Choices:
  tab_full <- dcast(tab_valid[dataname==name], model + generator ~ variable)
  col_names <- c(
    "Model",
    "Generator",
    measure_names
  )
  algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
  file_name <- sprintf(
    "paper/contents/table-%s-valid.tex", 
    data_indicator
  )
  cap <- sprintf(
    "All results for %s dataset: sample averages +/- one standard deviation over all valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
    name,
    data_indicator
  )
  kbl(
    tab_full, caption = cap, 
    align = "c", col.names=col_names, booktabs = T, escape=F, 
    format="latex"
  ) %>%
    kable_styling(latex_options = c("scale_down")) %>%
    kable_paper(full_width = F) %>%
    collapse_rows(columns = 1:2, latex_hline = "custom", valign = "top") %>%
    save_kable(file_name)
}
```

@@

```{r}
files <- list.files("artifacts/params/")
dt <- lapply(files[grepl(".csv", files)], function(x) {
    fread(file.path("artifacts/params/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
setcolorder(
  dt, 
  c(
    "dataname", "n_obs",
    "n_hidden", "n_layers", "activation", "n_ens",
    "epochs", "batch_size",
    "jem_sampling_steps", "sgld_batch_size", "lambda"
  )
)
dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
dt <- dt[order(dataname)]
dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
col_names <- c(
  "Dataset",
  "SGLD Steps", "Batch Size", "$\\lambda$"
)
kbl(
  dt_ebm, caption = "EBM hyperparemeter choices for our experiments. \\label{tab:ebmparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_ebm_params.tex")
```

@@

```{r}
dt_exp <- dt[,.(dataname, n_obs, n_hidden, n_layers, activation, n_ens, epochs, batch_size)]
col_names <- c(
  "Dataset", "Sample Size",
  "Hidden Units", "Hidden Layers", "Activation", "Ensemble Size",
  "Epochs", "Batch Size"
)
header <- c(" " = 2, "Network Architecture" = 4, "Training" = 2)
kbl(
  dt_exp, caption = "Paremeter choices for our experiments. \\label{tab:params} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  save_kable("paper/contents/table_params.tex")
```

```{r}
files <- list.files("artifacts/params/generator")
dt <- lapply(files, function(x) {
    fread(file.path("artifacts/params/generator", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt <- dt[,.(dataname,eta,λ1,λ3,λ3)]
dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
dt <- dt[order(dataname)]
col_names <- c(
  "Dataset",
  "$\\eta$", "$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$"
)
kbl(
  dt, caption = "Generator hyperparameters. \\label{tab:genparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_gen_params.tex")
```

```{r}
files <- list.files("artifacts/results/")
dt <- lapply(files[grepl("_model_performance.csv", files)], function(x) {
    fread(file.path("artifacts/results/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,dataname:=factor(dataname, levels=c("Linearly Separable", "Moons", "Circles", "MNIST", "GMSC"))]
dt <- dt[order(dataname,mod_name)]
setcolorder(
  dt, 
  c(
    "dataname", "mod_name",
    "acc", "precision", "f1score"
  )
)
col_names <- c("Dataset", "Model", "Accuracy", "Precision", "F1-Score")
kbl(
  dt, caption = "Various standard performance metrics for our different models grouped by dataset. \\label{tab:perf} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex", digits=2
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" "=2, "Performance Metrics" = 3)) %>%
  collapse_rows(columns = 1, latex_hline = "custom", valign = "top", custom_latex_hline = 1) %>%
  save_kable("paper/contents/table_perf.tex")
```

>>>>

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i[,variable:=factor(variable, levels=measures)]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,model:=factor(model,levels=chosen_model)]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep, table.env="table*"
) %>%
  kable_styling(
    latex_options = c("scale_down")
  ) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

@@

```{r}
for (name in unique(tab$dataname)) {
  data_indicator <- gsub(" ", "-", tolower(unique(name)))
  # Choices:
  tab_full <- dcast(tab[dataname==name], model + generator ~ variable)
  col_names <- c(
    "Model",
    "Generator",
    measure_names
  )
  algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
  file_name <- sprintf(
    "paper/contents/table-%s.tex", 
    data_indicator
  )
  cap <- sprintf(
    "All results for %s dataset: sample averages +/- one standard deviation over all counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}). \\label{tab:results-%s} \\newline",
    name,
    data_indicator
  )
  kbl(
    tab_full, caption = cap, 
    align = "c", col.names=col_names, booktabs = T, escape=F, 
    format="latex"
  ) %>%
    kable_styling(latex_options = c("scale_down")) %>%
    kable_paper(full_width = F) %>%
    collapse_rows(columns = 1:2, latex_hline = "custom", valign = "top") %>%
    save_kable(file_name)
}

@@

```{r}
for (name in unique(tab_valid$dataname)) {
  data_indicator <- gsub(" ", "-", tolower(unique(name)))
  # Choices:
  tab_full <- dcast(tab_valid[dataname==name], model + generator ~ variable)
  col_names <- c(
    "Model",
    "Generator",
    measure_names
  )
  algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
  file_name <- sprintf(
    "paper/contents/table-%s-valid.tex", 
    data_indicator
  )
  cap <- sprintf(
    "All results for %s dataset: sample averages +/- one standard deviation over all valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (\\textit{Wachter}). \\label{tab:results-%s} \\newline",
    name,
    data_indicator
  )
  kbl(
    tab_full, caption = cap, 
    align = "c", col.names=col_names, booktabs = T, escape=F, 
    format="latex"
  ) %>%
    kable_styling(latex_options = c("scale_down")) %>%
    kable_paper(full_width = F) %>%
    collapse_rows(columns = 1:2, latex_hline = "custom", valign = "top") %>%
    save_kable(file_name)
}
```

@@

```{r}
files <- list.files(paste0(res_path,"params"))
dt <- lapply(files[sapply(files, function(x) grepl("model_params.csv",x))], function(x) {
    fread(file.path(paste0(res_path,"params"), x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
setcolorder(
  dt, 
  c(
    "dataname", "n_obs",
    "n_hidden", "n_layers", "activation", "n_ens",
    "epochs", "batch_size",
    "jem_sampling_steps", "sgld_batch_size", "lambda"
  )
)
dataset_order = c(
  "Linearly Separable", "Moons", "Circles", 
  "California Housing", "GMSC", "German Credit",
  "MNIST", "Fashion MNIST"
)
dt[,dataname:=factor(dataname, levels=dataset_order)]
dt <- dt[order(dataname)]
dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
col_names <- c(
  "Dataset",
  "SGLD Steps", "Batch Size", "$\\lambda$"
)
kbl(
  dt_ebm, caption = "EBM hyperparemeter choices for our experiments. \\label{tab:ebmparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = ""
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_ebm_params.tex")
```

@@

```{r}
dt_exp <- dt[,.(dataname, n_obs, n_hidden, n_layers, activation, n_ens, epochs, batch_size)]
col_names <- c(
  "Dataset", "Sample Size",
  "Hidden Units", "Hidden Layers", "Activation", "Ensemble Size",
  "Epochs", "Batch Size"
)
header <- c(" " = 2, "Network Architecture" = 4, "Training" = 2)
kbl(
  dt_exp, caption = "Paremeter choices for our experiments. \\label{tab:params} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = ""
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  save_kable("paper/contents/table_params.tex")
```

```{r}
dt <- lapply(files[sapply(files, function(x) grepl("generator_params.csv",x))], function(x) {
    fread(file.path(paste0(res_path,"params"), x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt <- dt[,.(dataname,eta,lambda_1,lambda_2,lambda_3)]
dt[,dataname:=factor(dataname, levels=dataset_order)]
dt <- dt[order(dataname)]
col_names <- c(
  "Dataset",
  "$\\eta$", "$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$"
)
kbl(
  dt, caption = "Generator hyperparameters. \\label{tab:genparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = ""
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_gen_params.tex")
```

```{r}
dt <- lapply(files[grepl("_model_performance.csv", files)], function(x) {
    fread(file.path(paste0(res_path,"params"), x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,dataname:=factor(dataname, levels=dataset_order)]
dt <- dt[order(dataname,mod_name)]
setcolorder(
  dt, 
  c(
    "dataname", "mod_name",
    "acc", "precision", "f1score"
  )
)
col_names <- c("Dataset", "Model", "Accuracy", "Precision", "F1-Score")
kbl(
  dt, caption = "Various standard performance metrics for our different models grouped by dataset. \\label{tab:perf} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F,
  format="latex", digits=2, linesep = ""
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" "=2, "Performance Metrics" = 3)) %>%
  collapse_rows(columns = 1, latex_hline = "custom", valign = "top", custom_latex_hline = 1) %>%
  save_kable("paper/contents/table_perf.tex")
```
